Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference

### 2017-07-04 workflow/workflow.Rmd
merged my edits and Davide's edits

To infer lineages and pseudotimes, we use the package `slingshot`, with input the 4-dimensional MDS of the low-dimensional matrix `W`. We found that the slingshot results were robust to the number of dimensions _k_ for the MDS (we tried _k_ from 2 to 5). Here, we use the unsupervised version of `slingshot`, where we only provide the identity of the start cluster but not of the end clusters. This implementation recovers the lineages found in the published work (Figures \@ref(fig:curves) and \@ref(fig:tree)).

[SD: Refer to 3D version, used to create Figure 2.]

```{r}
our_cl <- primaryClusterNamed(ceObj)
cl = our_cl[our_cl != "-1"]
pal = pal[names(pal) != "-1"]
X <- W[our_cl != "-1", ]
mds <- cmdscale(dist(X), eig = TRUE, k = 4)
X <- mds$points

lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1")
lineages <- getCurves(lineages)
```

@@

```{r curves, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with smooth curves representing each inferred lineage."}
pairs(lineages, type="curves", pal[primaryClusterNamed(ceObj)])
```

```{r tree, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with connecting lines between cluster centers representing the inferred structure."}
pairs(lineages, type="lineages", col = pal[primaryClusterNamed(ceObj)])
```

@@

In the workflow, we recover a reasonable ordering of the clusters using the unsupervised version of slingshot. However, in some other cases, we have noticed that we need to give more guidance to the algorithm to find the correct ordering. `getLineages` has the option for the user to provide known end cluster(s). Here is the code to use `slingshot` in a supervised setting.

```{r slingshotsupervised, eval=FALSE}
lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1",
                         end.clus = c("c3", "c6"))
lineagees <- getCurves(lineages)
pairs(lineages, type="curves", pal[primaryClusterNamed(ceObj)])
pairs(lineages, type="lineages", col = pal[primaryClusterNamed(ceObj)])

@@

More formally, for each lineage, we use the robust local regression method `loess` to model in a flexible, non-linear manner the relationship between a gene's normalized expression measures and pseudotime. We then can test the null hypothesis of no change over time for each gene using the `gam` package. We implement this approach for the neuronal lineage and display the expression measures of the top 100 genes ordered by p-value in the heatmap of Figure \@ref(fig:heatmapsignificant).

>>>>

To infer lineages and pseudotimes, we use the package `slingshot`, with input the 4-dimensional MDS of the low-dimensional matrix `W`. We found that the slingshot results were robust to the number of dimensions _k_ for the MDS (we tried _k_ from 2 to 5). Here, we use the unsupervised version of `slingshot`, where we only provide the identity of the start cluster but not of the end clusters.

```{r}
our_cl <- primaryClusterNamed(ceObj)
cl = our_cl[our_cl != "-1"]
pal = pal[names(pal) != '-1']
X <- W[our_cl != "-1", ]
mds <- cmdscale(dist(X), eig = TRUE, k = 4)
X <- mds$points

lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1")
```

@@

Before fitting the simultaneous principal curves, we check the structure of the lineages by plotting the MST on the clusters. This shows that our implementation has recovered the lineages found in the published work (Figure \@ref(fig:tree)). The `slingshot` package also includes methods for 3-dimensional visualizations similar to \@ref(fig:stemcelldiff), using the `plot3d` function from the package `rgl`.

```{r tree, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with connecting lines between cluster centers representing the inferred structure."}
pairs(lineages, type='lineages', col = pal[primaryClusterNamed(ceObj)])
```

Having found the global lineage structure, we now construct a set of smooth, branching curves in order to infer the pseudotime variables. Simultaneous principal curves are constructed from the individual cells along each lineage, rather than the cell clusters. This makes them more stable and better suited for assigning cells to lineages. The final curves are shown in \@ref(fig:curves).

```{r curves, fig.cap="Cells color-coded by cluster in a 4-dimensional MDS space with smooth curves representing each inferred lineage."}
lineages <- getCurves(lineages)

pairs(lineages, type='curves', col = pal[primaryClusterNamed(ceObj)])
```

@@

In the workflow, we recover a reasonable ordering of the clusters using the unsupervised version of slingshot. However, in some other cases, we have noticed that we need to give more guidance to the algorithm to find the correct ordering. `getLineages` has the option for the user to provide known end cluster(s). Here is the code to use `slingshot` in a supervised setting, where we know that clusters `c8` and `c6` represent terminal cell fates.

```{r slingshotsupervised, eval=FALSE}
lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1",
                         end.clus = c("c3", "c6"))
lineagees <- getCurves(lineages)
pairs(lineages, type='curves', col = pal[primaryClusterNamed(ceObj)])
pairs(lineages, type='lineages', col = pal[primaryClusterNamed(ceObj)], show.constraints = TRUE)

@@

More formally, for each lineage, we use the robust local regression method loess to model in a flexible, non-linear manner the relationship between a gene's normalized expression measures and pseudotime. We then can test the null hypothesis of no change over time for each gene using the `gam` package. We implement this approach for the neuronal lineage and display the expression measures of the top 100 genes by p-value in the heatmap of Figure \@ref(fig:heatmapsignificant).
